On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-08-16 manuscripts/PNAS/chazthm_pnas.tex
k to 2k in Thm3 pf
sml updates

A vector $\mathbf{x}$ is said to be \emph{supported} in $S \subseteq [m]$ when $\mathbf{x} \in \text{\rm span}\{\mathbf{e}_j: j\in S\}$, where $\mathbf{e}_j$ are the standard basis in $\mathbb R^m$. 
%Denote by $\mathbf{x}^J$ the subvector formed from the entries of $\mathbf{x}$ indexed by $J$. 
A set of $k$-sparse vectors is said to be in \emph{general linear position} when any $k$ of them are linearly independent. The following is a precise statement of our main result.  For sake of brevity, the quantity $C_1 = C_1(\mathbf{A}, \{\mathbf{x}_i\}_{i=1}^N, \mathcal{H})$ is left undefined until Sec.~\ref{DUT} (see \eqref{Cdef1}).
%=======
%where we write $L_{k}$ when $\mathcal{H} = {[m] \choose k}$.\footnote{We note that $1 - \sqrt{k} L_k(\mathbf{M})$ is known as the asymmetric lower restricted isometry constant \cite{Blanchard2011}.} Clearly, for any $\mathbf{M}$ satisfying \eqref{SparkCondition}, we have $L_{k'}(\mathbf{M}) > 0$ for  $k' \leq 2k$. Note that if $\mathcal{H}$ covers $[m]$ (i.e., $\cup \mathcal{H} = [m]$), then $L_2 \geq L_{2\mathcal{H}}$.
% 
%A vector $\mathbf{x}$ is said to be \emph{supported} in $S \subseteq [m]$ when $\mathbf{x} \in \text{\rm span}\{\mathbf{e}_j: j\in S\}$, where $\mathbf{e}_j$ are the standard basis in $\mathbb R^m$. 
%%Denote by $\mathbf{x}^J$ the subvector formed from the entries of $\mathbf{x}$ indexed by $J$. 
%A set of $k$-sparse vectors is said to be in \emph{general linear position} when any $k$ of them are linearly independent. The following is a precise statement of our main result.  Only the quantity $C_1$ is left undefined until later (in Sec.~\ref{DUT}, Eq.~\eqref{Cdef1}).
%>>>>>>> d1407a275dc391dabad7bb62d3e659f9a4ac4624

@@

Another practical implication of Thm.\ref{DeterministicUniquenessTheorem} is the following: there is an effective procedure sufficient to affirm if a proposed solution to Prob.~\ref{InverseProblem} is indeed unique (up to noise and inherent ambiguities). One simply checks that the matrix and sparse codes satisfy the respective assumptions of Thm.~\ref{DeterministicUniquenessTheorem} on $\mathbf{A}$ and the $\mathbf{x}_i$.  
%Another practical implication of Thm.\ref{DeterministicUniquenessTheorem} is the following: there is an effective procedure sufficient to affirm if a proposed solution $(\mathbf{B}, \mathbf{\bar x}_1, \ldots, \mathbf{x}_N)$ to Prob.~\ref{InverseProblem} is indeed unique (up to noise and inherent ambiguities). One simply checks that $\mathbf{B}$ and the $\mathbf{\bar x}_i$ satisfy the respective assumptions of Thm.~\ref{DeterministicUniquenessTheorem}.  
%. on $\mathbf{A}$ and the $\mathbf{x}_i$, respectively.

@@

We furthermore note that unlike the requirements of previous works \cite{li2004analysis, Georgiev05, Aharon06, Hillar15}, the matrix $\mathbf{A}$ need not satisfy the spark condition to be recoverable from data. As a simple example for $k=2$, let $\mathbf{A} = [ \mathbf{e}_1, \ldots, \mathbf{e}_5, \mathbf{v}]$ where $\mathbf{v} = \mathbf{e}_1 + \mathbf{e}_3 + \mathbf{e}_5$, and take $\mathcal{H}$ to be all consecutive pairs of $[m]$ arranged in cyclic order. Then, $\mathbf{A}$ satisfies the assumptions of Thm.~\ref{DeterministicUniquenessTheorem} but not \eqref{SparkCondition}. %(since $\{ \mathbf{A}_1, \mathbf{A}_3, \mathbf{A}_5, \mathbf{A}_6\}$ is not a linearly independent set).
This weakening allows for a practical (polynomial) amount of data to still guarantee a stable dictionary in the case where the set of sparse supports for the $\mathbf{\bar x}_i$ is known to have a size that grows polynomially in $m, k$.

There are other less direct consequences of Thm.~\ref{DeterministicUniquenessTheorem}.  For instance, we use it to prove uniqueness for the optimization formulation of dictionary learning, Prob.~\ref{SLCopt}, the main object of interest for those applying dictionary learning to their data.

@@

Another extension of Thm.~\ref{DeterministicUniquenessTheorem} arises from the following analytic characterization of the spark condition.  Let $\mathbf{A}$  be the $n \times m$ matrix of $nm$ indeterminates $A_{ij}$. When real numbers are substituted for $A_{ij}$, the resulting matrix satisfies \eqref{SparkCondition} if and only if the following polynomial is nonzero:
\begin{align*}
f(\mathbf{A}) := \prod_{S \in {[m] \choose 2k}} \sum_{S' \in {[n] \choose 2k}} (\det \mathbf{A}_{S',S})^2,
\end{align*}
%
where for any $S' \in {[n] \choose 2k}$ and $S \in {[m] \choose 2k}$, the symbol $\mathbf{A}_{S',S}$ denotes the submatrix of entries $A_{ij}$ with $(i,j) \in S' \times S$.   We note that the large number of terms in this product is likely necessary due to the NP-hardness of deciding whether a given matrix $\mathbf{A}$ satisfies the spark condition \cite{tillmann2014computational}.

Since $f$ is analytic, having a single substitution of a real matrix $\mathbf{A}$ with $f(\mathbf{A}) \neq 0$ necessarily implies that the zeroes of $f$ form a set of measure zero. Fortunately, such a matrix $\mathbf{A}$ is easily constructed by adding rows of zeroes to any $\min(2k,m) \times m$ Vandermonde matrix as described above (so that each term in the product above for $f$ is nonzero). Hence, almost every $n \times m$ matrix with $n \geq \min(2k,m)$ satisfies \eqref{SparkCondition}.

@@

\begin{proposition}
Sparse neural population activity is recoverable from random (linear) noisy compression by any method that solves Prob.~\ref{DeterministicUniquenessTheorem} or \ref{SLCopt}; in particular, via biophysically plausible unsupervised sparse coding (e.g. \cite{rehnsommer2007, rozell2007neurally}).
\end{proposition}
%\cite{rehnsommer2007, rozell2007neurally, ganguli2012compressed, hu2014hebbian}

@@

The heart of the matter is therefore \eqref{Cstable}, which we now establish, first in the important special case $k = 1$

@@

In the field of theoretical neuroscience in particular, dictionary learning for sparse coding and related methods have recovered characteristic components of natural images \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent} and sounds \cite{bellsejnowski1996, smithlewicki2006, Carlson12} that some of the reproduce response properties of cortical neurons. Our results suggest that this correspondence could be due to the ``universality'' of sparse representations in natural data, an early idea in neural theory \cite{pitts1947}. %Furthermore, they justify the hypothesis of \cite{Isely10} that sparse codes passed through information bottlenecks in the brain are recovered from random projections via (unsupervised) biologically plausible sparse coding (e.g., \cite{rehnsommer2007, rozell2007neurally, ganguli2012compressed, hu2014hebbian}).
Deep learning noise issues: \cite{goodfellow2014explaining}.

@@

\begin{proof}[Proof (sketch) of Thm.~\ref{robustPolythm}]
Let $\textbf{M}$ be the matrix with columns $\mathbf{A}\mathbf{x}_i$, $i \in [N]$.  Consider the polynomial in the entries of $\mathbf{A}$ and $\mathbf{x}_i$:
\begin{align*}
g(\mathbf{A}, \{\mathbf{x}_i\}_{i=1}^N) := \prod_{S \in {[N] \choose k}} \sum_{S' \in {[n] \choose k}} (\det \textbf{M}_{S',S})^2,
\end{align*}
with notation as in Sec.~\ref{Results}.  
It can be checked that when $g$ is nonzero for a substitution of real numbers for the indeterminates, all of the genericity requirements on $\mathbf{A}$ and $\mathbf{x}_i$ in our proofs of stability in Thm.~\ref{DeterministicUniquenessTheorem} are satisfied (in particular, the spark condition on $\mathbf{A}$).
\end{proof}

>>>>

A vector $\mathbf{x}$ is said to be \emph{supported} in $S \subseteq [m]$ when $\mathbf{x} \in \text{\rm span}\{\mathbf{e}_j: j\in S\}$, where $\mathbf{e}_j$ are the standard basis in $\mathbb R^m$. 
%Denote by $\mathbf{x}^J$ the subvector formed from the entries of $\mathbf{x}$ indexed by $J$. 
A set of $k$-sparse vectors is said to be in \emph{general linear position} when any $k$ of them are linearly independent. The following is a precise statement of our main result.  The quantity $C_1 = C_1(\mathbf{A}, \{\mathbf{x}_i\}_{i=1}^N, \mathcal{H})$ is left undefined until Sec.~\ref{DUT} (see \eqref{Cdef1}).
%=======
%where we write $L_{k}$ when $\mathcal{H} = {[m] \choose k}$.\footnote{We note that $1 - \sqrt{k} L_k(\mathbf{M})$ is known as the asymmetric lower restricted isometry constant \cite{Blanchard2011}.} Clearly, for any $\mathbf{M}$ satisfying \eqref{SparkCondition}, we have $L_{k'}(\mathbf{M}) > 0$ for  $k' \leq 2k$. Note that if $\mathcal{H}$ covers $[m]$ (i.e., $\cup \mathcal{H} = [m]$), then $L_2 \geq L_{2\mathcal{H}}$.
% 
%A vector $\mathbf{x}$ is said to be \emph{supported} in $S \subseteq [m]$ when $\mathbf{x} \in \text{\rm span}\{\mathbf{e}_j: j\in S\}$, where $\mathbf{e}_j$ are the standard basis in $\mathbb R^m$. 
%%Denote by $\mathbf{x}^J$ the subvector formed from the entries of $\mathbf{x}$ indexed by $J$. 
%A set of $k$-sparse vectors is said to be in \emph{general linear position} when any $k$ of them are linearly independent. The following is a precise statement of our main result.  Only the quantity $C_1$ is left undefined until later (in Sec.~\ref{DUT}, Eq.~\eqref{Cdef1}).
%>>>>>>> d1407a275dc391dabad7bb62d3e659f9a4ac4624

@@

Another practical implication of Thm.\ref{DeterministicUniquenessTheorem} is the following: there is an effective procedure sufficient to affirm if a proposed solution to Prob.~\ref{InverseProblem} is indeed unique (up to noise and inherent ambiguities). One simply checks that the matrix and codes satisfy the respective assumptions of Thm.~\ref{DeterministicUniquenessTheorem} on $\mathbf{A}$ and the $\mathbf{x}_i$.  
%Another practical implication of Thm.\ref{DeterministicUniquenessTheorem} is the following: there is an effective procedure sufficient to affirm if a proposed solution $(\mathbf{B}, \mathbf{\bar x}_1, \ldots, \mathbf{x}_N)$ to Prob.~\ref{InverseProblem} is indeed unique (up to noise and inherent ambiguities). One simply checks that $\mathbf{B}$ and the $\mathbf{\bar x}_i$ satisfy the respective assumptions of Thm.~\ref{DeterministicUniquenessTheorem}.  
%. on $\mathbf{A}$ and the $\mathbf{x}_i$, respectively.

@@

We furthermore note that unlike the requirements of previous works \cite{li2004analysis, Georgiev05, Aharon06, Hillar15}, the matrix $\mathbf{A}$ need not satisfy the spark condition to be recoverable from data. As a simple example for $k=2$, let $\mathbf{A} = [ \mathbf{e}_1, \ldots, \mathbf{e}_5, \mathbf{v}]$ where $\mathbf{v} = \mathbf{e}_1 + \mathbf{e}_3 + \mathbf{e}_5$, and take $\mathcal{H}$ to be all consecutive pairs of $[m]$ arranged in cyclic order. Then, $\mathbf{A}$ satisfies the assumptions of Thm.~\ref{DeterministicUniquenessTheorem} but not \eqref{SparkCondition}. %(since $\{ \mathbf{A}_1, \mathbf{A}_3, \mathbf{A}_5, \mathbf{A}_6\}$ is not a linearly independent set).
This weakening allows for a practical (polynomial) amount of data to still guarantee a stable dictionary in the case where the set of sparse supports for the $\mathbf{\bar x}_i$ is known to have a size that grows polynomially in $m$ and $k$.

There are other less direct consequences of Thm.~\ref{DeterministicUniquenessTheorem}.  For instance, we use it to prove uniqueness for the optimization formulation of sparse linear coding, Prob.~\ref{SLCopt}, the main object of interest for those applying dictionary learning to their data.

@@

Another extension of Thm.~\ref{DeterministicUniquenessTheorem} arises from the following analytic characterization of the spark condition.  Let $\mathbf{A}$  be the $n \times m$ matrix of $nm$ indeterminates $A_{ij}$. When real numbers are substituted for $A_{ij}$, the resulting matrix satisfies condition \eqref{SparkCondition} if and only if the following polynomial is nonzero:
\begin{align*}
f(\mathbf{A}) := \prod_{S \in {[m] \choose 2k}} \sum_{S' \in {[n] \choose 2k}} (\det \mathbf{A}_{S',S})^2,
\end{align*}
%
where for any $S' \in {[n] \choose 2k}$ and $S \in {[m] \choose 2k}$, the symbol $\mathbf{A}_{S',S}$ denotes the submatrix of entries $A_{ij}$ with $(i,j) \in S' \times S$.   We note that the large number of terms in this product is likely necessary due to the NP-hardness of deciding whether a given matrix $\mathbf{A}$ satisfies the spark condition \cite{tillmann2014computational}.

Since $f$ is analytic, having a single substitution of a real matrix $\mathbf{A}$ with $f(\mathbf{A}) \neq 0$ necessarily implies that the zeroes of $f$ form a set of (Borel) measure zero. Fortunately, such a matrix $\mathbf{A}$ is easily constructed by adding rows of zeroes to any $\min(2k,m) \times m$ Vandermonde matrix as described above (so that each term in the product above for $f$ is nonzero). Hence, almost every $n \times m$ matrix with $n \geq \min(2k,m)$ satisfies \eqref{SparkCondition}.

@@

\begin{proposition}
Sparse neural population activity is recoverable from noisy random linear compression by any method that solves Prob.~\ref{DeterministicUniquenessTheorem} or \ref{SLCopt}; in particular, via biophysically plausible unsupervised sparse linear coding (e.g. \cite{rehnsommer2007, rozell2007neurally}).
\end{proposition}
%\cite{rehnsommer2007, rozell2007neurally, ganguli2012compressed, hu2014hebbian}

@@

The heart of the matter is therefore \eqref{Cstable}, which we now establish, first in the important special case of $k = 1$.

@@

In the field of theoretical neuroscience in particular, dictionary learning for sparse coding and related methods have recovered characteristic components of natural images \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent} and sounds \cite{bellsejnowski1996, smithlewicki2006, Carlson12} that some of the reproduce response properties of cortical neurons. Our results suggest that this correspondence could be due to the ``universality'' of sparse representations in natural data, an early idea in neural theory \cite{pitts1947}. %Furthermore, they justify the hypothesis of \cite{Isely10} that sparse codes passed through information bottlenecks in the brain are recovered from random projections via (unsupervised) biologically plausible sparse coding (e.g., \cite{rehnsommer2007, rozell2007neurally, ganguli2012compressed, hu2014hebbian}).
Deep learning noise issues: \cite{goodfellow2014explaining} and neuroscience theory \cite{ganguli2012compressed}.

@@

\begin{proof}[Proof (sketch) of Thm.~\ref{robustPolythm}]
Let $\textbf{M}$ be the matrix with columns $\mathbf{A}\mathbf{x}_i$, $i \in [N]$.  Consider the polynomial in the entries of $\mathbf{A}$ and $\mathbf{x}_i$:
\begin{align*}
g(\mathbf{A}, \{\mathbf{x}_i\}_{i=1}^N) := \prod_{S \in {[N] \choose 2k}} \sum_{S' \in {[n] \choose 2k}} (\det \textbf{M}_{S',S})^2,
\end{align*}
with notation as in Sec.~\ref{Results}.  
It can be checked that when $g$ is nonzero for a substitution of real numbers for the indeterminates, all of the genericity requirements on $\mathbf{A}$ and $\mathbf{x}_i$ in our proofs of stability in Thm.~\ref{DeterministicUniquenessTheorem} are satisfied (in particular, the spark condition on $\mathbf{A}$).
\end{proof}
